---
title: "Check in 4"
pagetitle: CheckIn4.qmd
author: "Theresa Szczepanski"
format:
  # pdf:
  #   toc: true
  #   number-sections: true
  #   colorlinks: true
  html:
    embed-resoures: true
    self-contained-math: true
    df-print: paged
    toc: true
    code-fold: true
    code-copy: true
    code-tools: true
editor: 
  markdown: 
    wrap: 72
---

# Research Question

I work for a public charter school in Massachusetts. Our students in
Grades 5-10 are tested annually in Mathematics, English Language Arts,
and Science. I am interested in using the student performance data to
identify areas of weakness in our program in terms of curricular
alignment. For every question on a given assessment, the state releases
an **item description** detailing what was asked of students and the
corresponding average score earned by students in our school as well as
the average score earned by students across Massachusetts.

***My research question is***: "Can test item descriptions on MCAS
assessments be mined to extract features associated with student
performance trends?"

# Hypothesis

I have already found statistically significant patterns in student
performance associated with an item's content reporting category in
every subject and grade level at our school.

In Mathematics, I have identified specific content reporting categories
that are relative weaknesses at different grade levels; however, this
does not take into account the differences in questions like those
Uurrutia and Araya classified in open-ended Math prompts. I would be
curious to see if our students are weaker in items that ask them to
evaluate an expression (apply technical skills in algebra/numeracy) vs.
construct or interpret a mathematical model (conceptual understanding).
This would be very interesting information for teachers.

When interviewing one of our experienced teachers who has historical
success with student achievement in English Language Arts (ELA), she
identified specific things that she believes all kids need to practice
for success with ELA, namely, "synthesizing multiple texts" and
"practice at the differences in reading informational text vs.
literature". These are requirements in questions that can be mined from
item descriptions but not from an item's reporting category or standard
description. Our students have historically performed weaker on the 7th
grade English Language Arts exam than on the 6th and 8th grade exams.
This suggests a curricular alignment issue. I've already identified
reporting categories in which our students have performed relatively
weaker on this assessment. I suspect that within these reporting
categories there exist patterns to the types of questions or tasks that
our students struggle with. This could provide valuable information for
teachers to adjust their instruction and instructional materials.

***Hypotheses***:

***H1***: A predictive model of student performance on Grade 5-10
Mathematics MCAS assessment items that includes regressors taken from
the test item descriptions will outperform a baseline predictive model
that includes only a given test item's content reporting category.

***H2***: A predictive model of student performance on Grade 5-10
English Language Arts MCAS assessment items that includes regressors
taken from the test item descriptions will outperform a baseline
predictive model that includes only a given test item's content
reporting category.

***H3***: A predictive model of student performance on Grade 5-8 grade
Science assessment items that includes regressors taken from the test
item descriptions will outperform a baseline predictive model that
includes only a given test item's content reporting category.

# Data Sources

I scraped the [Department of Elementary and Secondary Educations'
accountability
page](https://profiles.doe.mass.edu/mcas/mcasitems2.aspx?grade=07&subjectcode=ELA&linkid=10&orgcode=04830000&fycode=2023&orgtypecode=5&).
It includes tables for all Next Generation MCAS tests (beginning in
2018) and text descriptions for all test items. I have actually already
done this for the High School Science exams in my Python course last
fall. The structures of the tables are similar for the other exams.

Here is [my folder of Collab
notebooks](https://drive.google.com/drive/folders/1zCYUnbFfm-yqdpbM7R1O7iGeLj4kf1jU?usp=sharing)
with my Python code for scraping different grade level and subjects.
Here is a link to the notebook I used for the [HighSchool Science
MCAS](https://colab.research.google.com/drive/1loau1tMkYkROXcua6MMuYMs5NwzI_gv5?usp=drive_link)\

-   Math Corpus: 1010 Documents (item descriptions)

-   STE Corpus: 510 Documents, MS Only: 398 Documents

-   ELA Corpus: 693 Documents

In Uurrutia and Araya's paper, they used ***classification*** to
categorize open-ended mathematics questions into different types. I
would like to use ***dictionaries*** to classify MCAS questions into
different categories using the item descriptions.

Then I would like to use Supervised Machine Learning Methods that
include the features created from the dictionaries as well as features
that I extracted from the MCAS reporting tables to predict student
performance relative to their peers in the state, `RT-State Diff`, on a
given item.

I would like to use this same approach for the English Language Arts and
Science exams as well.

# Load Libraries

```{r}
library(caret)
library(devtools)
library(e1071)
library(ggplot2)
library(ggpubr)
library(hrbrthemes)
library(Metrics)
library(plotly)
library(plyr)
library(purrr)
library(quanteda)
library(quanteda.textstats)
library("quanteda.textplots")
library(randomForest)
library(RColorBrewer)
library(tidytext)
library(tidyverse)
library(tree)
library(viridis)



```

# Read-in/Tidy Data

Most of the cleaning was done by me in the process of scraping the
tables, which is included in [my folder of Python Collab
notebooks.](https://drive.google.com/drive/folders/1zCYUnbFfm-yqdpbM7R1O7iGeLj4kf1jU?usp=sharing)

## STE Data Frames

```{r}
STE_DF<-read_csv("STE_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `discipline core idea`, `standard`, `standard desc`, `pts`, `school%`, `state%`, school_state_diff, grade_level, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = as.factor(grade_level))

STE_DF
  

```

## ELA Data Frames

```{r}

ELA_G5_DF<-read_csv("G5_ELA_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `Cluster`, `text_type`, `standard`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 5)%>%
  mutate(grade_level = as.factor(grade_level))

#ELA_G5_DF


ELA_G6_DF<-read_csv("G6_ELA_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `Cluster`, `text_type`, `standard`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 6)%>%
  mutate(grade_level = as.factor(grade_level))

#ELA_G6_DF

ELA_G7_DF<-read_csv("G7_ELA_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `Cluster`, `text_type`, `standard`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 7)%>%
  mutate(grade_level = as.factor(grade_level))

#ELA_G7_DF

ELA_G8_DF<-read_csv("G8_ELA_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `Cluster`, `text_type`, `standard`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 8)%>%
  mutate(grade_level = as.factor(grade_level))

#ELA_G8_DF


ELA_G10_DF<-read_csv("G10_ELA_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `Cluster`, `text_type`, `standard`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 10)%>%
  mutate(grade_level = as.factor(grade_level))

#ELA_G10_DF

ELA_DF<-rbind(ELA_G5_DF, ELA_G6_DF)

ELA_DF<-rbind(ELA_DF, ELA_G7_DF)

ELA_DF<-rbind(ELA_DF, ELA_G8_DF)

ELA_DF<-rbind(ELA_DF, ELA_G10_DF)

ELA_DF<-ELA_DF%>%
  filter(`type` == "SR")

ELA_DF

```

## Math Data Frames

```{r}
Math_G5_DF<-read_csv("G5_Math_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `standard`, `Standard Description`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 5)%>%
  mutate(grade_level = as.factor(grade_level))

#Math_G5_DF


Math_G6_DF<-read_csv("G6_Math_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `standard`, `Standard Description`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 6)%>%
  mutate(grade_level = as.factor(grade_level))

#Math_G6_DF

Math_G7_DF<-read_csv("G7_Math_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `standard`, `Standard Description`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 7)%>%
  mutate(grade_level = as.factor(grade_level))

#Math_G7_DF

Math_G8_DF<-read_csv("G8_Math_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `standard`, `Standard Description`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 8)%>%
  mutate(grade_level = as.factor(grade_level))

#Math_G8_DF


Math_G10_DF<-read_csv("G10_Math_all_years_2024.csv")%>%
  select(`item description`, year, number, type, `reporting category`, `standard`, `Standard Description`,  `pts`, `school%`, `state%`, school_state_diff, item_library_URL)%>%
  mutate(number = as.factor(number))%>%
  mutate(grade_level = 10)%>%
  mutate(grade_level = as.factor(grade_level))

#Math_G10_DF

Math_DF<-rbind(Math_G5_DF, Math_G6_DF)

Math_DF<-rbind(Math_DF, Math_G7_DF)

Math_DF<-rbind(Math_DF, Math_G8_DF)

Math_DF<-rbind(Math_DF, Math_G10_DF)

Math_DF

```

# Methods/PreProcessing Approach

For each exam I followed a similar approach:

1.  Create a dataframe of all of the item descriptions and features
    across all grade levels and subjects. I kept as annotations all of
    the features provided by the state which include: `year`,
    `standard`, `reporting category`, `cluster` (sub category), `pts`,
    `state%`, `school%`, `school_state_diff`, and`type`(open response or
    selected response).

2.  Using `preText`, I looked for recommendations via the lowest score
    after running `factorial_preprocessing` and selected tokens
    accordingly. Interestingly, this recommended to not remove stop
    words in the Math and Science corpora. I believe this is perhaps
    because of the key logical role words like "and" ,"but", "not", and
    "or" play. These stop words do dominate the text descriptions
    though. Because of this, I also created a second set of tokens that
    included the removal of stopwords that I used for text
    visualizations and descriptive analysis of text. When applying
    dictionaries and using machine learning, I will use texts that
    include stop words. Also, whenever there was a recommendation to use
    stemming, I instead used lemmatization.

3.  Create subset document matrices for the items that students
    performed the **best** on and a matrix for the items that student
    performed the **worst** on.

4.  Create visuals of the word connections for all item descriptions as
    well as for the items descriptions in the **best** and **worst**
    subsets to explore differences.

5.  Using n-grams, identify key phrases from the item descriptions.

6.  Create a dictionary that uses key phrases identified in steps 4 and
    5 to tag an item as including any of the following

    Math:

    -   `modeling`: A modeling problem

    -   `apply_division`: Problem requiring the use of division and
        division facts

    -   `patterns_algebra`: Problem requiring the manipulation of
        equivalent algebraic or numeric expressions

    ELA:

    -   `poetry`: Students are asked questions involving where the
        source text is a poem

    -   `figurative_language`: question requiring the analysis of
        figurative language

    Science:

    -   `analyze graph`: Students are asked questions involving the
        analysis of a graph

    -   `analyze model`: Students are asked questions involving the
        analysis of a model

7.  Update the `Subject_DF` to include the new features identified via
    the subject area dictionary for each of the documents to use as a
    feature for Supervised Machine Learning model that will predict
    `school_state_diff`.
    
8. Create a Random Forest model using the features from the MCAS item reports and tune its parameters using 5-fold cross-validation.

9. Create a Random Forest model using the features from the MCAS and ***additional features*** extracted using dictionaries and the document feature matrices.

10. Compare the performance of the models as measured by %variance explained and MAE.

# Math Exam

## Create Corpus

```{r}
Math_item_corpus <-corpus(Math_DF, text_field = "item description")

#print(Math_item)

#(summary(Math_item_corpus))

Math_DF
```

## PreTextPreProcessing Decisions

Before completing the pre-processing process, I examined the different
choices using pre-Text. I was surprised to see that removing stop words
had a positive correlation coefficient.

I can see how key logical words like "not", "and", and "or", which are
also stop words can have a significant impact on the meaning of an exam
question. Perhaps, because each individual text is so small and the
texts are designed for assessing content skills and are not narrative
text, the stop words play more significant roles?

Given, these results, I will pre-process the data two ways, once using
the lowest recommended score, **N,** but also going lowercase as I can't
see how this has an effect on anything and has a correlation coefficient
of 0, and once using methods that includes removing stop words to use
for text visualization and simple descriptive summaries of the text:

-   Recommended preText score: "N-L" (remove punctuation and lowercase)

-   Alternative approach: "P-N-L-3" (remove punctuation, remove numbers,
    lower case, and n-grams)

![](MathPretext)

![](MathPretextCoef)

```{r}
# Sys.unsetenv("GITHUB_PAT")
# devtools::install_github("matthewjdenny/preText")
# library(preText)

```

```{r}

# preprocessed_documents_math <- factorial_preprocessing(
#     Math_item_corpus,
#     use_ngrams = TRUE,
#     infrequent_term_threshold = 0.2,
#     verbose = FALSE)
```

```{r}

# #names(preprocessed_documents_math)
# 
# head(preprocessed_documents_math$choices)
```

```{r}
# 
# preText_results <- preText(
#     preprocessed_documents_math,
#     dataset_name = "Math MCAS Item Descriptions",
#     distance_method = "cosine",
#     num_comparisons = 20,
#     verbose = FALSE)
# 
# 
# preText_score_plot(preText_results)
```

```{r}
# regression_coefficient_plot(preText_results,
#                             remove_intercept = TRUE)
```

### Tokenization 1: N-L

```{r}
## Extract the tokens

Math_item_tokens <- tokens(Math_item_corpus)

#print(Math_item_tokens)

```

```{r}

Math_item_tokens1 <- tokens(Math_item_corpus, 
    remove_numbers = T)


Math_item_tokens1 <- tokens_tolower(Math_item_tokens1)

# Math_item_tokens1 <- tokens_select(Math_item_tokens1,
#                    pattern = stopwords("en"),
#                   selection = "remove")

print(Math_item_tokens1)



```

### Tokenization 2: P-N-L-W & Lemmatization

```{r}
# remove punctuation and numbers
Math_item_tokens2 <- tokens(Math_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

# remove stopwords

Math_item_tokens2 <- tokens_select(Math_item_tokens2,
                   pattern = stopwords("en"),
                  selection = "remove")

# lower case

Math_item_tokens2 <- tokens_tolower(Math_item_tokens2)




print(Math_item_tokens2)
```

### lemmatization

```{r}
lem_Math_item_tokens2<-tokens_replace(Math_item_tokens2,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

lem_Math_item_tokens2

```

## Create DFM

```{r}

df_Math_toks1<-dfm(Math_item_tokens1)


df_Math_toks1_smaller<- dfm_trim(df_Math_toks1, min_docfreq = 0.08, docfreq_type = "prop")


df_Math_toks1

df_Math_toks1_smaller


df_Math_toks2<-dfm(lem_Math_item_tokens2)


df_Math_toks_smaller<- dfm_trim(df_Math_toks2, min_docfreq = 0.08, docfreq_type = "prop")


df_Math_toks2

df_Math_toks_smaller

```

### n-grams

```{r}
 Sys.unsetenv("GITHUB_PAT")
devtools::install_github("slanglab/phrasemachine/R/phrasemachine")
#   
library(phrasemachine) 


```

```{r}

documents_weak_items<-as.character(corpus_subset(Math_item_corpus, school_state_diff < -2))[26:29]

#documents_weak_items

phrases<-phrasemachine(documents_weak_items, minimum_ngram_length = 2,
                         maximum_ngram_length = 4,
                         return_phrase_vectors = TRUE,
                         return_tag_sequences = TRUE)

# look at some example phrases
print(phrases[[1]]$phrases[1:20])

```

```{r}
documents_weak_items

```

## Subset: Worst performing Items

```{r}
Math_worst_item_corpus<- corpus_subset(Math_item_corpus, school_state_diff < 2 )

# remove punctuation and numbers
Math_worst_tokens <- tokens(Math_worst_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

Math_worst_tokens <- tokens_tolower(Math_worst_tokens)


Math_worst_tokens <-  tokens_select(Math_worst_tokens,
                  pattern = stopwords("en"),
                 selection = "remove")

lem_Math_worst_tokens<-tokens_replace(Math_worst_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_Math_worst_tokens<-dfm(lem_Math_worst_tokens)
df_Math_worst_toks_smaller<- dfm_trim(df_Math_worst_tokens, min_docfreq = 0.08, docfreq_type = "prop")

```

## Subset: Best performing Items

```{r}
Math_best_item_corpus<- corpus_subset(Math_item_corpus, school_state_diff > 5 )

# remove punctuation and numbers
Math_best_tokens <- tokens(Math_best_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

Math_best_tokens <- tokens_tolower(Math_best_tokens)


Math_best_tokens <-  tokens_select(Math_best_tokens,
                  pattern = stopwords("en"),
                 selection = "remove")

lem_Math_best_tokens<-tokens_replace(Math_best_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_Math_best_tokens<-dfm(lem_Math_best_tokens)
df_Math_best_toks_smaller<- dfm_trim(df_Math_best_tokens, min_docfreq = 0.08, docfreq_type = "prop")

```

## Descriptive Analysis

Interestingly, some words like "determine", "real-world", and
"expression" are top features from the Math tokens as a whole ***but***,
in the tokens taken only from the worst performing items, the word
expression appears ***much*** more frequently than the phrase
"real-world", and this is the opposite in our best performing items and
the entire corpus. Also, the word "graph" is a top-feature in the Math
corpus and the best performing items but not a top-feature in the worst
performing items.

## All Items: Top Features

```{r}
topfeatures(df_Math_toks2, 20)

#topfeatures(df_Math_toks1, 20)

```

## Worstperforming Items: Top Features

```{r}

topfeatures(df_Math_worst_toks_smaller, 20)


```

## Best performing Items: Top Features

```{r}
topfeatures(df_Math_best_toks_smaller, 20)
```

## Visualizations

I found it particularly interesting when I saw the connections on the
worst performing subset vs. the best performing subset. There appears to
be a theme of identifying equivalent expressions that only appears in
the worst subset. Solving "real-world" problems; i.e. Modeling, appears
strongly in ***both*** our best and worst subset. This makes me curious
if there are certain mathematical topics where our students are
struggling with the modeling problems.

Because of these patterns, I will create a dictionary to classify
identify if an item requires `Modeling`, `apply_division`, or
`patterns_algebra`.

### Word Cloud: All Questions

```{r}
#dfm_item
smaller_dfm <-dfm_trim(df_Math_toks2, min_termfreq = 10)
smaller_dfm <- dfm_trim(smaller_dfm, min_docfreq = 0.02, docfreq_type = "prop")

textplot_wordcloud(smaller_dfm)

```

### All questions

Clearly, mathematical modeling is a significant skill throughout the
Math exam as the network "solve real-world problem" and the connection
"real-world context" are very strong.

The word "expression"

```{r}
### Word connections for all items

dim(df_Math_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_Math_toks_smaller))

# create plot
textplot_network(df_Math_toks_smaller, vertex_size = size/ max(size) * 3)
```

### Worst-performing questions Network Plot

The connection "equivalent expression" and the connections to the word
"expression" are ***much stronger*** in the network of students' weakest
performing items. Noticeable, the words "context" and "represent" are
not present as strong connectors to the word "expression" however they
are in the best items network. This signals to me that that the
algebraic manipulation of expressions is a weakness rather than
constructing an expression in a mathematical modeling context.

The network "solve real-world problem" is also strong in our weakest
performing item descriptions but not as heavily emphasized as in our
strongest performing items.

```{r}
### Word connections for worst items

dim(df_Math_worst_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_Math_worst_toks_smaller))

# create plot
textplot_network(df_Math_worst_toks_smaller, vertex_size = size/ max(size) * 3)
```

### Best-performing questions

We can see the clear thread of "solve real-world problems" is strongly
present in our students' best performing items' descriptions.

Also, I can see the connection "determine expression" present as well as
"represent" and "context". These strike me as words connected with
constructing models and identifying mathematical concepts in word
problems.

```{r}
### Word connections for best items

dim(df_Math_best_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_Math_best_toks_smaller))

# create plot
textplot_network(df_Math_best_toks_smaller, vertex_size = size/ max(size) * 3)
```

## Dictionary to Create New Features

## Dictionary: Key Math Terms:

```{r}
my_math_dict <- dictionary(list(modeling=c("real-world","context", "word_problem"),             
              patterns_alg=c("equivalent expression", "distributive", "identify equivalent", "distribute", "factor", "parentheses"),
              apply_division=c("divide","division","quotient", "quotients")))

my_math_dict

# patterns_alg=c("equivalent expression","radical expression", "rational expression", #"expression", "eqivalent", "distributive", "distribute", "factor")
```

```{r}
math_Toks1_mydict <- df_Math_toks1 %>%
  dfm_lookup(my_math_dict)

tail(math_Toks1_mydict, 10)

df_MathDict <- convert(math_Toks1_mydict, to = "data.frame")



df_MathDict

#Math_DF<-left_join(df_MathDict, Math_DF, by = "doc_id")

```

```{r}
# mathToksDFM_mydict <- df_Math_worst_tokens %>%
#   dfm_lookup(my_math_dict)
# 
# head(mathToksDFM_mydict, 10)
# 
# summary(df_Math_toks1)

```

```{r}
Math_DF<-mutate(Math_DF, ID = as.character(row_number()))

Math_DF<-mutate(Math_DF, doc_id = str_trim(paste("text", ID, sep="")))

Math_DF<-left_join(df_MathDict, Math_DF, by = "doc_id")

```

```{r}
Math_DF<-Math_DF%>%
  mutate(modeling_factor = case_when(
    modeling == 0 ~ "no",
    modeling > 0 ~ "yes"
  ))%>%
  mutate(patterns_alg_factor = case_when(
    patterns_alg == 0 ~ "no",
    patterns_alg > 0 ~ "yes"
  ))%>%
  mutate(apply_division_factor = case_when(
    apply_division == 0 ~ "no",
    apply_division > 0 ~ "yes"
  ))

  

```

## Modeling Visualization

```{r}

Math_DF%>%
  #filter(modeling > 0)%>%
  filter(grade_level == 10)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`modeling_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G10 MCAS: Student Performance Modeling by Reporting Category")

```

### G5 Modeling

```{r}

Math_DF%>%
  #filter(modeling > 0)%>%
  filter(grade_level == 5)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`modeling_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G5 MCAS: Student Performance Modeling by Reporting Category")

Math_DF%>%
  filter(grade_level == 5)%>%
  filter(modeling_factor == "yes")%>%
  filter(`reporting category` == "GE")
  

```

### G6 Modeling

```{r}

Math_DF%>%
  #filter(modeling > 0)%>%
  filter(grade_level == 6)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`modeling_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G6 MCAS: Student Performance Modeling by Reporting Category")

#Math_DF
Math_DF%>%
  filter(grade_level == 6)%>%
  filter(modeling_factor == "yes")%>%
  filter(`reporting category` == "RP")%>%
  select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
  arrange(`school_state_diff`)


Math_DF%>%
  filter(grade_level == 6)%>%
  filter(modeling_factor == "yes")%>%
  filter(`reporting category` == "GE")%>%
  select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
  arrange(`school_state_diff`)


# Math_DF%>%
#   filter(grade_level == 6)%>%
#   filter(modeling_factor == "yes")%>%
#   filter(`reporting category` == "SP")%>%
#   select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
#   arrange(`school_state_diff`)
#   

```

### G7 Modeling

```{r}

Math_DF%>%
  #filter(modeling > 0)%>%
  filter(grade_level == 7)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`modeling_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G7 MCAS: Student Performance Modeling by Reporting Category")

Math_DF
Math_DF%>%
  filter(grade_level == 7)%>%
  filter(modeling_factor == "yes")%>%
  filter(`reporting category` == "EE")%>%
  select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
  arrange(`school_state_diff`)


```

### G8 Modeling

```{r}

Math_DF%>%
  #filter(modeling > 0)%>%
  filter(grade_level == 8)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`modeling_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G8 MCAS: Student Performance Modeling by Reporting Category")


Math_DF%>%
  filter(grade_level == 8)%>%
  filter(modeling_factor == "no")%>%
  filter(`reporting category` == "SP")%>%
  select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
  arrange(`school_state_diff`)


```

## Patterns_Alg Visual

The concept of equivalent expressions, the distributive property, and
factoring seems to be emphasized and particularly challenging for our
students in grades 6, 7, and 10. It is not heavily emphasized on the G8
exam.

```{r}
Math_DF%>%
  #filter(patterns_alg > 0)%>%
  filter(grade_level == 6 | grade_level == 7 | grade_level == 10)%>%
  ggplot( aes(x=`patterns_alg_factor`, y=`school_state_diff`, fill=`patterns_alg_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G6, G7, G10 MCAS: Student Performance Patterns Algebra")#+
 #facet_wrap(~`reporting category`)


```

### G6

```{r}

Math_DF%>%
  #filter(patterns_alg > 0)%>%
  filter(grade_level == 6)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`patterns_alg_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G6 MCAS: Student Performance Patterns Algebra and Reporting Category")


Math_DF%>%
  filter(grade_level == 6)%>%
  filter(patterns_alg_factor == "yes")%>%
 # filter(`reporting category` == "SP")%>%
  select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
  arrange(`school_state_diff`)



```

## Apply Division Visual

### G6

G6 is the when students are expected to be fluent in the standard
algorithm for long division. Applying division seems to have been a
factor in student success in some G6 reporting categories

```{r}
Math_DF%>%
  #filter(patterns_alg > 0)%>%
  filter(grade_level == 6)%>%
  ggplot( aes(x=`apply_division_factor`, y=`school_state_diff`, fill=`apply_division_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G6 MCAS: Student Performance Apply Division")#+
 #facet_wrap(~`grade_level`)


```

```{r}
Math_DF%>%
  filter(grade_level == 6)%>%
  filter(`reporting category` == "NS" | `reporting category` == "EE")%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`apply_division_factor`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("Math G6 MCAS: Student Performance Apply_Division and Reporting Category")




Math_DF%>%
  filter(grade_level == 6)%>%
  filter(apply_division_factor == "yes")%>%
  filter(patterns_alg_factor == "no")%>%
 # filter(`reporting category` == "SP")%>%
  select(`standard`, `item description`, `school_state_diff`, `year`, item_library_URL)%>%
  arrange(`school_state_diff`)



```

## Machine Learning Model to Predict school_state_diff

### Create Training Set

```{r}
#Math_DF
Math_DF_Train_Full<-Math_DF%>%
  mutate(year = as.factor(year))%>%
  mutate(`school_state_bool` = case_when(
    `school_state_diff` <= 0 ~ "negative",
    `school_state_diff` > 0 ~ "positive")
  )%>%
  #filter(grade_level == 5)%>%
  filter(grade_level == 10 | grade_level == 8| grade_level == 7| grade_level == 6|grade_level == 5)%>%
 mutate(id = row_number())%>%
  select(`doc_id`, `id`, `standard`, `reporting category`, `type`, `year`,  `modeling_factor`, `patterns_alg_factor`, `apply_division_factor`, `state%`, `school_state_bool`, `school_state_diff`, `grade_level`)

Math_DF_Train_Full

```

### Create Training Set

```{r}
# set seed
set.seed(12345)

N<-nrow(Math_DF_Train_Full)

trainIndex <- sample(1:N, .8*N)

testIndex<-c(1:N)[-trainIndex]

# check length of training set
length(trainIndex)

trainIndexDF<-trainIndex

#Math_DF_Test<-Math_DF_Train_Full[-Math_DF_Train]


trainIndexDF<-as.data.frame(t(trainIndex))
dim(trainIndexDF)
#dim(Math_DF_Test)

#Create training vector

trainIndexDF

trainIndexDF<-trainIndexDF%>%
  pivot_longer(cols = starts_with("v"), names_to = "value_id", values_to = "id")
#trainIndexDF<-mutate(trainIndexDF, doc_id = str_trim(paste("text", ID, #sep="")))%>%
#  select("doc_id")

trainIndexDF

Math_DF_Train_Full

Math_DF_Train<-left_join(trainIndexDF, Math_DF_Train_Full, by = "id")

Math_DF_Train




```

### Create Test Set

```{r}
testIndexDF<-testIndex

#Math_DF_Test<-Math_DF_Test_Full[-Math_DF_Test]


testIndexDF<-as.data.frame(t(testIndex))
dim(testIndexDF)
#dim(Math_DF_Test)

#Create testing vector

#testIndexDF

testIndexDF<-testIndexDF%>%
  pivot_longer(cols = starts_with("v"), names_to = "value_id", values_to = "id")
#testIndexDF<-mutate(testIndexDF, doc_id = str_trim(paste("text", ID, sep="")))%>%
#  select("doc_id")

#Math_DF_Test_Full

Math_DF_Test<-left_join(testIndexDF, Math_DF_Train_Full, by = "id")



```

### Random Forest Model

#### Base Regression

## Cross Validation

### Random forest with text features

```{r}
data_ctrl<-trainControl(method = "cv", number = 5)

Math_DF_Train_Full_Text<-Math_DF_Train_Full%>%
  select(`standard`, `reporting category`, `type`, `year`, `modeling_factor`, `state%`, `grade_level`, `school_state_diff` )

Math_DF_Train_Full_Base<-Math_DF_Train_Full%>%
  select(`standard`, `reporting category`, `type`, `year`, `state%`, `grade_level`, `school_state_diff` )
```

```{r}
model_caret_rf_text<-train(school_state_diff ~.
                      , data = Math_DF_Train_Full_Text,
                      trControl = data_ctrl,
                      method = "rf")

```

```{r}
model_caret_rf_text

varImpPlot(model_caret_rf_text$finalModel)


```

```{r}
model_caret_rf<-train(school_state_diff ~.
                      , data = Math_DF_Train_Full_Base,
                      trControl = data_ctrl,
                      method = "rf")

```

```{r}
model_caret_rf

```

```{r}

Math_DF_Train_X<-Math_DF_Train%>%
  select( `standard`, `reporting category`, `type`, `year`,  `state%`, grade_level)



Math_DF_Train_Y<-Math_DF_Train%>%
  select(`school_state_bool`)

Math_DF_Train_Y

Math_DF_Test_X<-Math_DF_Test%>%
  select( `standard`, `reporting category`, `type`, `year`,  `state%`, grade_level)

Math_DF_Test_Y<-Math_DF_Test%>%
  select(`school_state_diff`)


#Math_DF_Test_Y
#Math_DF_Train_Y

#Math_DF_Train_X

#Math_DF_Test_X



```

```{r}
set.seed(444)

diff_RF<-randomForest(Math_DF_Train_X,
                      y = Math_DF_Train$school_state_diff,
                      xtest = Math_DF_Test_X,
                      ytest = Math_DF_Test$school_state_diff,
                      mtry =103,
                      importance = TRUE,
                      ntree = 1010,
                      type="regression")

diff_RF

```

#### Regression w/ text

```{r}
Math_DF_Train_X<-Math_DF_Train%>%
  select( `standard`, `reporting category`, `type`, `year`, `state%`, `modeling_factor`, `patterns_alg_factor`, grade_level)



Math_DF_Train_Y<-Math_DF_Train%>%
  select(`school_state_diff`)

Math_DF_Train_Y

Math_DF_Test_X<-Math_DF_Test%>%
  select( `standard`, `reporting category`, `type`, `year`, `state%`, `modeling_factor`, `patterns_alg_factor`, grade_level)

Math_DF_Test_Y<-Math_DF_Test%>%
  select(`school_state_diff`)

```

```{r}
set.seed(444)

diff_RF_text<-randomForest(Math_DF_Train_X,
                      y = Math_DF_Train$school_state_diff,
                      xtest = Math_DF_Test_X,
                      ytest = Math_DF_Test$school_state_diff,
                      mtry = 2,
                      importance = TRUE,
                      ntree = 1010,
                      type="regression")

```

```{r}

diff_RF_text

```

```{r}
varImpPlot(diff_RF_text)


```

### Support Vector Regression

***to do*** address the issue that new standards appear in the test set
and thus the svm can't make a prediction

```{r}
set.seed(446)

Math_SVM_Train<-Math_DF_Train%>%
  select(`standard`, `reporting category`, `type`, `year`, `state%`, `modeling_factor`, `patterns_alg_factor`, `apply_division_factor`, grade_level, school_state_diff)
#Math_DF_Train_X
#Math_DF_Train$school_state_diff

svm_text<-svm(school_state_diff ~., data = Math_SVM_Train)
                     
svm_text

#svm_pred<-predict(svm_text, Math_DF_Test_X)

```

# Science, Technology, and Engineering Exam (STE) Exam

## Create Corpus

```{r}
STE_item_corpus <-corpus(STE_DF, text_field = "item description")

#print(STE_item)

#summary(STE_item_corpus)

#STE_DF
```

## PreTextPreProcessing Decisions

Before completing the pre-processing process, I examined the different
choices using pre-Text. I was surprised to see that removing stopwords
had a positive correlation coefficient; yet the combination "P-W" had
the lowest score of the Pre-text results. I can see how key logical
words like "not", "and", and "or", which are also stop words can have a
significant impact on the meaning of an exam question. Perhaps, because
each individual text is so small and the texts are designed for
assessing content skills and are not narrative text, the stop words play
more significant roles?

Given, these results, I will pre-process the data two ways, once using
the lowest recommended score, **P-W** and once using methods that
includes, lowercase and lemmatization of words to use for creating
visualizations

-   Recommended preText score: "P-W" (remove punctuation and stopwords)

-   Alternative approach: "P-N-L-W" + Lemmatization (remove punctuation,
    remove numbers, lower case, remove stopwords and lemmatization)

    ![](STEPretext)

![](STEPretextCoef)

```{r}
# Sys.unsetenv("GITHUB_PAT")
# devtools::install_github("matthewjdenny/preText")
# library(preText)

```

```{r}
# 
# preprocessed_documents_STE <- factorial_preprocessing(
#     STE_item_corpus,
#     use_ngrams = TRUE,
#     infrequent_term_threshold = 0.2,
#     verbose = FALSE)
```

```{r}

#names(preprocessed_documents_STE)

#head(preprocessed_documents_STE$choices)
```

```{r}

# preText_results <- preText(
#     preprocessed_documents_STE,
#     dataset_name = "STE MCAS Item Descriptions",
#     distance_method = "cosine",
#     num_comparisons = 20,
#     verbose = FALSE)
# 
# 
# preText_score_plot(preText_results)
```

```{r}
# regression_coefficient_plot(preText_results,
#                             remove_intercept = TRUE)
```

### Tokenization 1: P-W

```{r}
## Extract the tokens

STE_item_tokens <- tokens(STE_item_corpus)

#print(STE_item_tokens)

```

```{r}

STE_item_tokens1 <- tokens(STE_item_corpus, 
    remove_punct = T)

STE_item_tokens1 <- tokens_select(STE_item_tokens1,
                   pattern = stopwords("en"),
                  selection = "remove")

print(STE_item_tokens1)



```

### Tokenization 2: P-N-L-W- Lemmatization

```{r}
# remove punctuation and numbers
STE_item_tokens2 <- tokens(STE_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

STE_item_tokens2 <- tokens_select(STE_item_tokens2,
                   pattern = stopwords("en"),
                  selection = "remove")

# lower case

STE_item_tokens2 <- tokens_tolower(STE_item_tokens2)


# remove stopwords

#print(STE_item_tokens2)
```

### lemmatization

When I originally made word clouds, I noticed object and objects
appearing separately as well as model and models. I believe these are
important, so I chose to also do lemmatization for creating visual summaries.

```{r}
lem_STE_item_tokens2<-tokens_replace(STE_item_tokens2,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

lem_STE_item_tokens2


```

## Create DFMs

### n-grams

When I examine the phrases in the first ten documents of weak items, the weakest documents seem to consistently involve the analysis of graphs/diagrams/models. I would like to capture this as a text feature.

```{r}
library(phrasemachine) 
documents_weak_items<-as.character(corpus_subset(STE_item_corpus, school_state_diff < -2))[1:10]

#documents_weak_items

phrases<-phrasemachine(documents_weak_items, minimum_ngram_length = 2,
                         maximum_ngram_length = 4,
                         return_phrase_vectors = TRUE,
                         return_tag_sequences = TRUE)

# look at some example phrases
print(phrases[[1]]$phrases[1:20])

```

```{r}
documents_weak_items

```

```{r}
df_STE_toks2<-dfm(lem_STE_item_tokens2)

df_STE_toks1<-dfm(STE_item_tokens1)



```

#### Worst performing Items

```{r}

STE_worst_item_corpus<- corpus_subset(STE_item_corpus, school_state_diff < 2 )

# remove punctuation and numbers
STE_worst_tokens <- tokens(STE_worst_item_corpus, 
    remove_punct = T,
    remove_numbers = T)



STE_worst_tokens<-STE_worst_tokens <-tokens_tolower(STE_worst_tokens)

STE_worst_tokens<- tokens_select(STE_worst_tokens, 
                     pattern = stopwords("en"),
                     selection = "remove")

lem_STE_worst_tokens<-tokens_replace(STE_worst_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_STE_worst_toks<-dfm(lem_STE_worst_tokens)

df_STE_worst_toks_smaller<-dfm_trim(df_STE_worst_toks, min_docfreq = .05, docfreq_type = "prop")

```

### MS only

#### Subset: Worst Items

```{r}
STE_MS_item_corpus<- corpus_subset(STE_item_corpus, grade_level != 9 )

#summary(STE_MS_item_corpus)


STE_MS_worst_item_corpus<-corpus_subset(STE_MS_item_corpus, school_state_diff < 2)

# remove punctuation and numbers
STE_MS_worst_tokens <- tokens(STE_MS_worst_item_corpus, 
    remove_punct = T,
    remove_numbers = T)


STE_MS_worst_tokens<-STE_worst_tokens <-tokens_tolower(STE_MS_worst_tokens)

STE_MS_worst_tokens<- tokens_select(STE_MS_worst_tokens, 
                     pattern = stopwords("en"),
                     selection = "remove")


lem_STE_MS_worst_tokens<-tokens_replace(STE_MS_worst_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)



df_STE_MS_worst_toks<-dfm(lem_STE_MS_worst_tokens)

df_STE_MS_worst_toks_smaller<-dfm_trim(df_STE_worst_toks, min_docfreq = .05, docfreq_type = "prop")

```

#### Subset: Best Items

```{r}

STE_MS_best_item_corpus<-corpus_subset(STE_MS_item_corpus, school_state_diff > 5)


# remove punctuation and numbers
STE_MS_best_tokens <- tokens(STE_MS_best_item_corpus, 
    remove_punct = T,
    remove_numbers = T)


STE_MS_best_tokens<-STE_best_tokens <-tokens_tolower(STE_MS_best_tokens)

STE_MS_best_tokens<- tokens_select(STE_MS_best_tokens, 
                     pattern = stopwords("en"),
                     selection = "remove")


lem_STE_MS_best_tokens<-tokens_replace(STE_MS_best_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)



df_STE_MS_best_toks<-dfm(lem_STE_MS_best_tokens)

df_STE_MS_best_toks_smaller<-dfm_trim(df_STE_MS_best_toks, min_docfreq = .06, docfreq_type = "prop")

#textplot_wordcloud(df_STE_MS_best_toks)
#textplot_wordcloud(df_STE_MS_best_toks_smaller)

```

## Descriptive Analysis

When I examine the top features, it is notable that the word ***graph*** appears as a top feature of the worst items but not in the best performing items. The words ***analyze*** and ***model*** appear in both the worst and best performing top features; however, words that are associated with physical sciences, like ***force*** appear in the worst items and words associated with life sciences, like ***organism***, ***plant***, and ***population*** appear only in the best performing top features.

## All Items: Top Features

```{r}
topfeatures(df_STE_toks2, 20)

#topfeatures(df_Math_toks1, 20)

```

## Worstperforming Items: Top Features

```{r}
STE_worst_item_corpus<- corpus_subset(STE_item_corpus, school_state_diff < 0 )

# remove punctuation and numbers
STE_worst_tokens <- tokens(STE_worst_item_corpus, 
    remove_punct = T,
    remove_numbers = T)



STE_worst_tokens<-STE_worst_tokens <-tokens_tolower(STE_worst_tokens)

STE_worst_tokens<- tokens_select(STE_worst_tokens, 
                     pattern = stopwords("en"),
                     selection = "remove")

lem_STE_worst_tokens<-tokens_replace(STE_worst_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_STE_worst_toks<-dfm(lem_STE_worst_tokens)

df_STE_worst_toks_smaller<-dfm_trim(df_STE_worst_toks, min_docfreq = .06, docfreq_type = "prop")
```

```{r}

topfeatures(df_STE_worst_toks_smaller, 20)


```


## Best performing Items: Top Features

```{r}
STE_best_item_corpus<- corpus_subset(STE_item_corpus, school_state_diff > 2 )

# remove punctuation and numbers
STE_best_tokens <- tokens(STE_best_item_corpus, 
    remove_punct = T,
    remove_numbers = T)



STE_best_tokens<-STE_best_tokens <-tokens_tolower(STE_best_tokens)

STE_best_tokens<- tokens_select(STE_best_tokens, 
                     pattern = stopwords("en"),
                     selection = "remove")


lem_STE_best_tokens<-tokens_replace(STE_best_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_STE_best_toks<-dfm(lem_STE_best_tokens)

df_STE_best_toks_smaller<-dfm_trim(df_STE_best_toks, min_docfreq = .08, docfreq_type = "prop")

```

```{r}
topfeatures(df_STE_best_toks_smaller, 20)
```
## MS Worst: Top Features
```{r}

topfeatures(df_STE_MS_worst_toks_smaller, 20)

```


## MS Best: Top Features
```{r}
topfeatures(df_STE_MS_best_toks_smaller, 20)

```

## Visualizations

It was evident that the worst item tokens seem to largely represent text
of items in the physical sciences. Our 9th grade students take a science
exam that is only on Physics and as you can see below, they have
performed weaker against their peers in the state compared to Rising
Tide students at other grade levels. It would therefore, be worth
exploring the worst items for Middle School (5th-8th grade) students.
When I look at the Mifdle School student performance by Reporting
Category, one of their weakest categories appears to be Physical
Sciences. So this is also contributing to the dominance of Physical
Science terms in the worst items tokens.

```{r}
#STE_DF

STE_DF%>%
  filter(grade_level != 9)%>%
  group_by(`reporting category`)%>%
  summarize(mean_diff = mean(school_state_diff))

STE_DF%>%
  #filter(grade_level != 9)%>%
  #filter(Course == "STEM"| Course == "Math" & `Reporting Category` == "Geometry")%>%
  #filter(`Cluster` != "Convert like measurement units within a given measurement system.")%>%
  #filter(`Cluster` != "Geometric measurement: Understand concepts of volume and relate volume to multiplication and to addition.")%>%
  #filter(`year` == 2024 | year == 2019)%>%
  ggplot( aes(x=`grade_level`, y=`school_state_diff`, fill=`grade_level`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("STE MCAS: Student Performance by Grade Level") +
    xlab("")#+



STE_DF%>%
  filter(grade_level != 9)%>%
  #filter(`year` == 2024 | year == 2019)%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`reporting category`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("MS STE MCAS: Student Performance by Reporting Category") +
    xlab("")#+


STE_DF%>%
  filter(grade_level != 9)%>%
  #filter(Course == "STEM"| Course == "Math" & `Reporting Category` == "Geometry")%>%
  filter(`reporting category` == "PS")%>%
  #filter(`Cluster` != "Geometric measurement: Understand concepts of volume and relate volume to multiplication and to addition.")%>%
  #filter(`year` == 2024 | year == 2019)%>%
  ggplot( aes(x=`discipline core idea`, y=`school_state_diff`, fill=`discipline core idea`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
     # legend.position="none",
      axis.text.x=element_blank(),
      plot.title = element_text(size=11)
    ) +
    ggtitle("MS STE MCAS: Student Performance in Physical Sciences") +
    xlab("")#+


  


```

### Word Cloud

The word cloud has Physical science words larger than other subjects.
This also suggests a need to explore the Middle School separately as the
5-8th grade exams give equal weight to Physical Sciences, Life Science,
Earth Science, and Technology and Engineering.

```{r}


#dfm_item


textplot_wordcloud(df_STE_toks2)

smaller_dfm <-dfm_trim(df_STE_toks2, min_termfreq = 10)
smaller_dfm <- dfm_trim(smaller_dfm, min_docfreq = 0.07, docfreq_type = "prop")

textplot_wordcloud(smaller_dfm, min_count =3, random_order = FALSE)
```

### Examine text visuals on worst-performing questions

```{r}

textplot_wordcloud(df_STE_worst_toks)
textplot_wordcloud(df_STE_worst_toks_smaller)

```

### Examine text visual on best-performing questions

```{r}


textplot_wordcloud(df_STE_best_toks)
textplot_wordcloud(df_STE_best_toks_smaller)

```

### Word Connections

#### Worst items

```{r}

dim(df_STE_worst_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_STE_worst_toks_smaller))

# create plot
textplot_network(df_STE_worst_toks_smaller, vertex_size = size/ max(size) * 3)
```

#### Best Items

```{r}

# check the dimensions (i.e., the number of rows and the number of columnns) of the matrix we created

dim(df_STE_best_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_STE_best_toks_smaller))

# create plot
textplot_network(df_STE_best_toks_smaller, vertex_size = size/ max(size) * 3)




```

## Word connections for all items

```{r}

# check the dimensions (i.e., the number of rows and the number of columnns) of the matrix we created
smaller_fcm <- fcm(smaller_dfm)
dim(df_STE_best_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(smaller_fcm))

# create plot
textplot_network(smaller_dfm, vertex_size = size/ max(size) * 3)
```

## MS Worst

```{r}


# check the dimensions (i.e., the number of rows and the number of columnns) of the matrix we created
#smaller_fcm <- fcm(smaller_dfm)
dim(df_STE_best_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(smaller_fcm))

# create plot
textplot_network(smaller_dfm, vertex_size = size/ max(size) * 3)

```

## MS Best

```{r}
#df_STE_MS_best_toks_smaller

dim(df_STE_MS_best_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_STE_MS_best_toks_smaller))

# create plot
textplot_network(df_STE_MS_best_toks_smaller, vertex_size = size/ max(size) * 3)

```

## Dictionary to Create New Features

## Dictionary: Key STE Terms:

```{r}
my_ste_dict <- dictionary(list(model=c("Analyze force model", "Analyze particle models", "Analyze a graph", "Analyze a diagram", "Interpret a graph", "model", "models", "diagram", "diagram", "graph", "graphs"),             
              algebra=c("calculate")))

my_ste_dict

# patterns_alg=c("equivalent expression","radical expression", "rational expression", #"expression", "eqivalent", "distributive", "distribute", "factor")
```

```{r}
STE_Toks1_mydict <- df_STE_toks1 %>%
  dfm_lookup(my_ste_dict)

tail(STE_Toks1_mydict, 10)

df_STEDict <- convert(STE_Toks1_mydict, to = "data.frame")



df_STEDict

#Math_DF<-left_join(df_MathDict, Math_DF, by = "doc_id")

```

```{r}
# mathToksDFM_mydict <- df_Math_worst_tokens %>%
#   dfm_lookup(my_math_dict)
# 
# head(mathToksDFM_mydict, 10)
# 
# summary(df_Math_toks1)

```

```{r}
STE_DF<-mutate(STE_DF, ID = as.character(row_number()))

STE_DF<-mutate(STE_DF, doc_id = str_trim(paste("text", ID, sep="")))

STE_DF<-left_join(df_STEDict, STE_DF, by = "doc_id")

STE_DF

```

```{r}
STE_DF<-STE_DF%>%
  mutate(algebra_factor = case_when(
    algebra == 0 ~ "no",
    algebra > 0 ~ "yes"
  ))%>%
  mutate(model_factor = case_when(
    model == 0 ~ "no",
    model > 0 ~ "yes"
  ))

  

```

## Machine Learning Training Set

```{r}
STE_DF
STE_DF_Train_Full<-STE_DF%>%
   mutate(year = as.factor(year))%>%
#   mutate(`school_state_bool` = case_when(
#     `school_state_diff` <= 0 ~ "negative",
#     `school_state_diff` > 0 ~ "positive")
#   )%>%
#  filter(grade_level == 10)%>%
#   filter(grade_level == 10 | grade_level == 8| grade_level == 7| grade_level == 6|grade_level == 5)%>%
  mutate(id = row_number())%>%
   select(`id`, `standard`, `reporting category`, `discipline core idea`, `pts`, `year`, `type`, `state%`, `school_state_diff`, `grade_level`, model_factor, algebra_factor)
# 
 STE_DF_Train_Full

```

### Create Training Set
```{r}
# set seed
set.seed(12345)

N<-nrow(STE_DF_Train_Full)

trainIndex <- sample(1:N, .8*N)

testIndex<-c(1:N)[-trainIndex]

# check length of training set
length(trainIndex)

trainIndexDF<-trainIndex

#Math_DF_Test<-Math_DF_Train_Full[-Math_DF_Train]


trainIndexDF<-as.data.frame(t(trainIndex))
dim(trainIndexDF)
#dim(Math_DF_Test)

#Create training vector

trainIndexDF

trainIndexDF<-trainIndexDF%>%
  pivot_longer(cols = starts_with("v"), names_to = "value_id", values_to = "id")
#trainIndexDF<-mutate(trainIndexDF, doc_id = str_trim(paste("text", ID, #sep="")))%>%
#  select("doc_id")

#trainIndexDF

STE_DF_Train_Full

STE_DF_Train<-left_join(trainIndexDF, STE_DF_Train_Full, by = "id")

STE_DF_Train

STE_DF_Train_X<-STE_DF_Train%>%
  select(`standard`, `reporting category`, `discipline core idea`, `pts`, `year`, `type`, `state%`, `grade_level`, model_factor, algebra_factor)

STE_DF_Train_Base_X<-STE_DF_Train%>%
  select(`standard`, `reporting category`, `discipline core idea`, `pts`, `year`, `type`, `state%`, `grade_level`)


STE_DF_Train_Y<-STE_DF_Train$school_state_diff

```

### Create Test Set

```{r}
testIndexDF<-testIndex

#STE_DF_Test<-STE_DF_Test_Full[-STE_DF_Test]


testIndexDF<-as.data.frame(t(testIndex))
#dim(testIndexDF)
#dim(STE_DF_Test)

#Create testing vector

#testIndexDF

testIndexDF<-testIndexDF%>%
  pivot_longer(cols = starts_with("v"), names_to = "value_id", values_to = "id")
#testIndexDF<-mutate(testIndexDF, doc_id = str_trim(paste("text", ID, sep="")))%>%
#  select("doc_id")

#STE_DF_Test_Full

STE_DF_Test<-left_join(testIndexDF, STE_DF_Train_Full, by = "id")

STE_DF_Test_X<-STE_DF_Test%>%
select(`standard`, `reporting category`, `discipline core idea`, `pts`, `year`, `type`, `state%`, `grade_level`, model_factor, algebra_factor)

STE_DF_Test_Base_X<-STE_DF_Test%>%
select(`standard`, `reporting category`, `discipline core idea`, `pts`, `year`, `type`, `state%`, `grade_level`)


STE_DF_Test_Y<-STE_DF_Test%>%
  select(`school_state_diff`)


STE_DF_Test_Y
STE_DF_Train_Y

STE_DF_Train_X

STE_DF_Test_X
STE_DF_Test_Base_X

```

### Cross Validation

### Tune:Random Forest w/ text features

```{r}
data_ctrl<-trainControl(method = "cv", number = 5)

STE_DF_Train_Full_Text<-STE_DF_Train_Full%>%
  select(`standard`, `reporting category`, `type`, `year`, `discipline core idea`, `pts`, `model_factor`, `algebra_factor`, `state%`, `grade_level`, `school_state_diff` )

STE_DF_Train_Full_Base<-STE_DF_Train_Full%>%
  select(`standard`, `reporting category`, `type`, `year`, `discipline core idea`, `pts`, `state%`, `grade_level`, `school_state_diff` )
```

```{r}
ste_model_caret_rf_text<-train(school_state_diff ~.
                      , data = STE_DF_Train_Full_Text,
                      trControl = data_ctrl,
                      method = "rf")

```

```{r}
ste_model_caret_rf_text

ste_model_caret_rf_text$finalModel


```

```{r}
ste_model_caret_rf<-train(school_state_diff ~.
                      , data = STE_DF_Train_Full_Base,
                      trControl = data_ctrl,
                      method = "rf")

```

```{r}
ste_model_caret_rf$finalModel

```





### Random Forest Model

#### Regression Base Model

```{r}
set.seed(444)

STE_diff_RF<-randomForest(STE_DF_Train_Base_X,
                      y = STE_DF_Train$school_state_diff,
                      xtest = STE_DF_Test_Base_X,
                      ytest = STE_DF_Test$school_state_diff,
                      mtry = 3,
                      importance = TRUE,
                      type="regression",
                      ntree = 500)


STE_diff_RF
```



```{r}

varImpPlot(STE_diff_RF)

```







### Random Forest Model w/ text features

#### Regression

```{r}
set.seed(444)

STE_diff_RF_Text<-randomForest(STE_DF_Train_X,
                      y = STE_DF_Train$school_state_diff,
                      xtest = STE_DF_Test_X,
                      ytest = STE_DF_Test$school_state_diff,
                      mtry = 4,
                      importance = TRUE,
                      type="regression",
                      ntree = 500)


STE_diff_RF_Text
```

```{r}
varImpPlot(STE_diff_RF_Text)

```

# English Language Arts Exam

## Create Corpus

```{r}
ELA_item_corpus <-corpus(ELA_DF, text_field = "item description")

#print(ELA_item)

summary(ELA_item_corpus)

#ELA_DF
```

## PreText PreProcessing Decisions

Before completing the pre-processing process, I examined the different
choices using pre-Text. I was surprised to see that removing stopwords
had a positive correlation coefficient; yet the combination "P-W" had
the lowest score of the Pre-text results. I can see how key logical
words like "not", "and", and "or", which are also stop words can have a
significant impact on the meaning of an exam question. Perhaps, because
each individual text is so small and the texts are designed for
assessing content skills and are not narrative text, the stop words play
more significant roles?

Given, these results, I will pre-process the data two ways, once using
the lowest recommended score and once including lower case and lemmatization to use for visualzations.

-   Recommended preText score: "S-W-3" (remove stop words,
    Lemmatization/Stemming)

-   Alternative approach: "P-N-W-L" and Lemmatization (remove
    punctuation, remove numbers, lower case, and lemmatization)

![](ELAPretextCoef)

Also, due to teacher input I am also creating a subcorpus of items that
assessed Literature and items that assessed Informational text. Teachers
describe having to change their approach to teaching these things and
students having different weaknesses according to text type.

```{r}
# Sys.unsetenv("GITHUB_PAT")
# devtools::install_github("matthewjdenny/preText")
# library(preText)

```

```{r}

# preprocessed_documents_ELA <- factorial_preprocessing(
#     ELA_item_corpus,
#     use_ngrams = TRUE,
#     infrequent_term_threshold = 0.2,
#     verbose = FALSE)
```

```{r}

# names(preprocessed_documents_ELA)
# 
# head(preprocessed_documents_ELA$choices)
```

```{r}

# preText_results <- preText(
#     preprocessed_documents_ELA,
#     dataset_name = "ELA MCAS Item Descriptions",
#     distance_method = "cosine",
#     num_comparisons = 20,
#     verbose = FALSE)
# 
# 
# preText_score_plot(preText_results)
```

```{r}
# regression_coefficient_plot(preText_results,
#                             remove_intercept = TRUE)
```

### Tokenization 1: L-W-Lemma


```{r}
## Extract the tokens

ELA_item_tokens <- tokens(ELA_item_corpus)

print(ELA_item_tokens)

```

```{r}

ELA_item_tokens1 <- tokens_tolower(ELA_item_tokens)

ELA_item_tokens1 <- tokens_select(ELA_item_tokens1,
                   pattern = stopwords("en"),
                  selection = "remove")

print(ELA_item_tokens1)



```

### Tokenization 2: P-N-L-W & Lemmatization

```{r}
# remove punctuation and numbers
ELA_item_tokens2 <- tokens(ELA_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

# remove stopwords

ELA_item_tokens2 <- tokens_select(ELA_item_tokens2,
                   pattern = stopwords("en"),
                  selection = "remove")

# lower case

ELA_item_tokens2 <- tokens_tolower(ELA_item_tokens2)




print(ELA_item_tokens2)
```

### lemmatization

```{r}
lem_ELA_item_tokens2<-tokens_replace(ELA_item_tokens2,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

lem_ELA_item_tokens2

lem_ELA_item_tokens1<-tokens_replace(ELA_item_tokens1,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

lem_ELA_item_tokens1


```

## Create DFMs

```{r}
df_ELA_toks2<-dfm(lem_ELA_item_tokens2)

df_ELA_toks1<-dfm(lem_ELA_item_tokens1)

```


### n-grams

When I examine the phrases in the first ten documents of weak items, the weakest documents seem to consistently involve the analysis of graphs/diagrams/models. I would like to capture this as a text feature.

```{r}
library(phrasemachine) 
documents_weak_items<-as.character(corpus_subset(ELA_item_corpus, school_state_diff < -2))[10:15]

#documents_weak_items

phrases<-phrasemachine(documents_weak_items, minimum_ngram_length = 2,
                         maximum_ngram_length = 4,
                         return_phrase_vectors = TRUE,
                         return_tag_sequences = TRUE)

# look at some example phrases
print(phrases[[1]]$phrases[1:20])

```

```{r}
documents_weak_items

```

## Worst performing Items

### Literature

```{r}
ELA_worst_lit_item_corpus<- corpus_subset(ELA_item_corpus, school_state_diff < 2 & text_type == "Literature" )

# remove punctuation and numbers
ELA_worst_lit_tokens <- tokens(ELA_worst_lit_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

ELA_worst_lit_tokens <- tokens_tolower(ELA_worst_lit_tokens)


ELA_worst_lit_tokens <-  tokens_select(ELA_worst_lit_tokens,
                   pattern = stopwords("en"),
                  selection = "remove")

lem_ELA_worst_lit_tokens<-tokens_replace(ELA_worst_lit_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_ELA_worst_lit_tokens<-dfm(lem_ELA_worst_lit_tokens)
df_ELA_worst_lit_toks_smaller<- dfm_trim(df_ELA_worst_lit_tokens, min_docfreq = 0.09, docfreq_type = "prop")

```

### Informational

```{r}
ELA_worst_inf_item_corpus<- corpus_subset(ELA_item_corpus, school_state_diff < 2 & text_type == "Informational" )

# remove punctuation and numbers
ELA_worst_inf_tokens <- tokens(ELA_worst_inf_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

ELA_worst_inf_tokens <- tokens_tolower(ELA_worst_inf_tokens)


ELA_worst_inf_tokens <-  tokens_select(ELA_worst_inf_tokens,
                   pattern = stopwords("en"),
                  selection = "remove")

lem_ELA_worst_inf_tokens<-tokens_replace(ELA_worst_inf_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_ELA_worst_inf_tokens<-dfm(lem_ELA_worst_inf_tokens)
df_ELA_worst_inf_toks_smaller<- dfm_trim(df_ELA_worst_inf_tokens, min_docfreq = 0.09, docfreq_type = "prop")

```

## Best performing Items

### Literature

```{r}
ELA_best_lit_item_corpus<- corpus_subset(ELA_item_corpus, school_state_diff > 5 & text_type == "Literature" )

# remove punctuation and numbers
ELA_best_lit_tokens <- tokens(ELA_best_lit_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

ELA_best_lit_tokens <- tokens_tolower(ELA_best_lit_tokens)


ELA_best_lit_tokens <-  tokens_select(ELA_best_lit_tokens,
                   pattern = stopwords("en"),
                  selection = "remove")

lem_ELA_best_lit_tokens<-tokens_replace(ELA_best_lit_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_ELA_best_lit_tokens<-dfm(lem_ELA_best_lit_tokens)
df_ELA_best_lit_toks_smaller<- dfm_trim(df_ELA_best_lit_tokens, min_docfreq = 0.09, docfreq_type = "prop")

```

### Informational

```{r}
ELA_best_inf_item_corpus<- corpus_subset(ELA_item_corpus, school_state_diff > 5 & text_type == "Informational" )

# remove punctuation and numbers
ELA_best_inf_tokens <- tokens(ELA_best_inf_item_corpus, 
    remove_punct = T,
    remove_numbers = T)

ELA_best_inf_tokens <- tokens_tolower(ELA_best_inf_tokens)


ELA_best_inf_tokens <-  tokens_select(ELA_best_inf_tokens,
                   pattern = stopwords("en"),
                  selection = "remove")

lem_ELA_best_inf_tokens<-tokens_replace(ELA_best_inf_tokens,
                             pattern = lexicon:: hash_lemmas$token,
                             replacement = lexicon:: hash_lemmas$lemma)

df_ELA_best_inf_tokens<-dfm(lem_ELA_best_inf_tokens)
df_ELA_best_inf_toks_smaller<- dfm_trim(df_ELA_best_inf_tokens, min_docfreq = 0.09, docfreq_type = "prop")

```

## Descriptive Analysis

When I examine the top features, it is notable that the words ***evidence*** and ***support*** appear as top features in informational text but not in literature. The word ***poem*** appears much higher and more frequently as a top feature of in ***worst literature items****.

## All Items: Top Features

```{r}
topfeatures(df_ELA_toks2, 20)

#topfeatures(df_Math_toks1, 20)

```

## Lit: Worstperforming Items: Top Features

```{r}

topfeatures(df_ELA_worst_lit_toks_smaller, 20)


```


## Lit Best performing Items: Top Features

```{r}
topfeatures(df_ELA_best_lit_toks_smaller, 20)
```
## Inf: Worstperforming Items: Top Features

```{r}
topfeatures(df_ELA_worst_inf_toks_smaller, 20)

```


## Inf Best performing Items: Top Features

```{r}

topfeatures(df_ELA_best_inf_toks_smaller, 20)

```



## Visualizations

```{r}
#ELA_DF

ELA_DF%>%
  group_by(`reporting category`)%>%
  summarize(mean_diff = mean(school_state_diff))


ELA_DF%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`text_type`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle(" ELA MCAS: Student Performance by Reporting Category") 


ELA_DF%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`Cluster`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("ELA MCAS: Student Performance by Reporting Category")
  

ELA_DF%>%
  filter(text_type == "Literature")%>%
  ggplot( aes(x=`reporting category`, y=`school_state_diff`, fill=`Cluster`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11)
    ) +
    ggtitle("ELA MCAS: Student Performance by Reporting Category") +
    labs(caption = "Literature") 




ELA_DF%>%
  filter(`reporting category` == "LA")%>%
  ggplot( aes(x=`Cluster`, y=`school_state_diff`, fill=`text_type`)) +
    geom_boxplot() +
    scale_fill_viridis(discrete = TRUE, alpha=0.6) +
    geom_jitter(color="black", size=0.4, alpha=0.9) +
    theme_ipsum() +
    theme(
      #legend.position="none",
      plot.title = element_text(size=11),
      axis.text.x = element_text(angle = 45, hjust = 1)
    ) +
    ggtitle("ELA MCAS: Student Performance by Language Cluster and Text Type") 




```

### Word Cloud

```{r}
#dfm_item
textplot_wordcloud(df_ELA_toks2)

smaller_dfm <-dfm_trim(df_ELA_toks2, min_termfreq = 5)
smaller_dfm <- dfm_trim(smaller_dfm, min_docfreq = 0.05, docfreq_type = "prop")

textplot_wordcloud(smaller_dfm, min_count =3, random_order = FALSE)
```

### Word Connections

#### Literature

##### Best Items

```{r}
### Word connections for worst items

dim(df_ELA_best_lit_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_ELA_best_lit_toks_smaller))

# create plot
textplot_network(df_ELA_best_lit_toks_smaller, vertex_size = size/ max(size) * 3)
```

##### Worst Items

```{r}
### Word connections for worst items

dim(df_ELA_worst_lit_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_ELA_worst_lit_toks_smaller))

# create plot
textplot_network(df_ELA_worst_lit_toks_smaller, vertex_size = size/ max(size) * 3)
```

### Word Connections

#### Informational

##### Visual Best Items

```{r}
### Word connections for worst items

dim(df_ELA_best_inf_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_ELA_best_inf_toks_smaller))

# create plot
textplot_network(df_ELA_best_inf_toks_smaller, vertex_size = size/ max(size) * 3)
```

##### Worst Items

```{r}
### Word connections for best items

dim(df_ELA_worst_inf_toks_smaller)

# compute size weight for vertices in network
size <- log(colSums(df_ELA_worst_inf_toks_smaller))

# create plot
textplot_network(df_ELA_worst_inf_toks_smaller, vertex_size = size/ max(size) * 3)
```

## Dictionary to Create New Features

## Dictionary: Key ELA Terms:

```{r}
my_ela_dict <- dictionary(list(poetry=c("poem","stanza", "lyric", "verse"),
                               evidence = c("claim", "support", "evidence", "supports", "purpose"),
                               fig_language = c("figurative language", "simile", "metaphor")))

my_ela_dict


```

```{r}
ELA_Toks1_mydict <- df_ELA_toks1 %>%
  dfm_lookup(my_ela_dict)

tail(ELA_Toks1_mydict, 10)

df_ELADict <- convert(ELA_Toks1_mydict, to = "data.frame")



df_ELADict

#Math_DF<-left_join(df_MathDict, Math_DF, by = "doc_id")

```



```{r}
ELA_DF<-mutate(ELA_DF, ID = as.character(row_number()))

ELA_DF<-mutate(ELA_DF, doc_id = str_trim(paste("text", ID, sep="")))

ELA_DF<-left_join(df_ELADict, ELA_DF, by = "doc_id")

ELA_DF

```

```{r}
ELA_DF<-ELA_DF%>%
  mutate(poetry_factor = case_when(
    poetry == 0 ~ "no",
    poetry > 0 ~ "yes"
  ))%>%
    mutate(fig_language_factor = case_when(
    fig_language == 0 ~ "no",
    fig_language > 0 ~ "yes"
  ))%>%
    
    mutate(evidence_factor = case_when(
    evidence == 0 ~ "no",
    evidence > 0 ~ "yes"
  ))
    

  

```

## Machine Learning

```{r}
ELA_DF
ELA_DF_Train_Full<-ELA_DF%>%
   mutate(year = as.factor(year))%>%
#   mutate(`school_state_bool` = case_when(
#     `school_state_diff` <= 0 ~ "negative",
#     `school_state_diff` > 0 ~ "positive")
#   )%>%
  filter(grade_level == 10| grade_level ==8)%>%
  # filter(grade_level == 10 | grade_level == 8| grade_level == 7| grade_level == 6|grade_level == 5)%>%
  mutate(id = row_number())%>%
   select(`id`, `standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`, `school_state_diff`, `grade_level`, `poetry_factor`, fig_language_factor, evidence_factor)
# 
 ELA_DF_Train_Full

```

### Create Training Set

```{r}
# set seed
set.seed(12345)

N<-nrow(ELA_DF_Train_Full)

trainIndex <- sample(1:N, .8*N)

testIndex<-c(1:N)[-trainIndex]

# check length of training set
#length(trainIndex)

trainIndexDF<-trainIndex

#Math_DF_Test<-Math_DF_Train_Full[-Math_DF_Train]


trainIndexDF<-as.data.frame(t(trainIndex))
dim(trainIndexDF)
#dim(Math_DF_Test)

#Create training vector

#trainIndexDF

trainIndexDF<-trainIndexDF%>%
  pivot_longer(cols = starts_with("v"), names_to = "value_id", values_to = "id")
#trainIndexDF<-mutate(trainIndexDF, doc_id = str_trim(paste("text", ID, #sep="")))%>%
#  select("doc_id")

#trainIndexDF

#ELA_DF_Train_Full

ELA_DF_Train<-left_join(trainIndexDF, ELA_DF_Train_Full, by = "id")

ELA_DF_Train

ELA_DF_Train_X<-ELA_DF_Train%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`,  `grade_level`, `poetry_factor`, fig_language_factor, evidence_factor)

ELA_DF_Train_Base_X<-ELA_DF_Train%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`,  `grade_level`)


ELA_DF_Train_Y<-ELA_DF_Train$school_state_diff

```

### Create Test Set

```{r}
testIndexDF<-testIndex

#ELA_DF_Test<-ELA_DF_Test_Full[-ELA_DF_Test]


testIndexDF<-as.data.frame(t(testIndex))
#dim(testIndexDF)
#dim(ELA_DF_Test)

#Create testing vector

#testIndexDF

testIndexDF<-testIndexDF%>%
  pivot_longer(cols = starts_with("v"), names_to = "value_id", values_to = "id")
#testIndexDF<-mutate(testIndexDF, doc_id = str_trim(paste("text", ID, sep="")))%>%
#  select("doc_id")

#ELA_DF_Test_Full

ELA_DF_Test<-left_join(testIndexDF, ELA_DF_Train_Full, by = "id")

ELA_DF_Test_X<-ELA_DF_Test%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`,  `grade_level`, `poetry_factor`, fig_language_factor, evidence_factor)

ELA_DF_Test_Base_X<-ELA_DF_Test%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`,  `grade_level`)


ELA_DF_Test_Y<-ELA_DF_Test$school_state_diff





```

### Cross Validation

### Tune:Random Forest w/ text features

```{r}
data_ctrl<-trainControl(method = "cv", number = 5)

ELA_DF_Train_Full_Text<-ELA_DF_Train_Full%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`, `school_state_diff`, `grade_level`, `poetry_factor`, fig_language_factor, evidence_factor, `school_state_diff` )

ELA_DF_Train_Full_Base<-ELA_DF_Train_Full%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`, `school_state_diff`, `grade_level`, `school_state_diff` )
```

```{r}
ela_model_caret_rf_text<-train(school_state_diff ~.
                      , data = ELA_DF_Train_Full_Text,
                      trControl = data_ctrl,
                      method = "rf")

```

```{r}
ela_model_caret_rf_text

ela_model_caret_rf_text$finalModel


```


### Tune:Base Random Forest
```{r}
ela_model_caret_rf<-train(school_state_diff ~.
                      , data = ELA_DF_Train_Full_Base,
                      trControl = data_ctrl,
                      method = "rf")

```

```{r}

ela_model_caret_rf
ela_model_caret_rf$finalModel

```




```{r}
set.seed(444)

ELA_diff_RF<-randomForest(ELA_DF_Train_Base_X,
                      y = ELA_DF_Train$school_state_diff,
                      xtest = ELA_DF_Test_Base_X,
                      ytest = ELA_DF_Test$school_state_diff,
                      mtry = 2,
                      importance = TRUE,
                      type="regression",
                      ntree = 500)


```

```{r}
ELA_diff_RF
#predicted<-predict(ELA_diff_RF, ELA_DF_Test_X)
#mae(predicted, ELA_DF_Test$school_state_diff)

#actual <- ELA_DF_Test$school_state_diff




#R2 <- 1 - (sum((actual-predicted)^2)/sum((actual-mean(actual))^2))

#R2

# mean(abs(actual-predicted))
# 
# mean(abs(actual))
# sd(abs(actual))
# sd(abs(actual-predicted))

```

```{r}

varImpPlot(ELA_diff_RF)

```

### RF Regression w/text



```{r}
set.seed(444)

ELA_diff_RF_text<-randomForest(x = ELA_DF_Train_X,
                      y = ELA_DF_Train$school_state_diff,
                      xtest = ELA_DF_Test_X,
                      ytest = ELA_DF_Test$school_state_diff,
                      mtry = 2,
                      importance = TRUE,
                      type="regression",
                      ntree = 500)


```

```{r}
ELA_diff_RF_text
#predicted<-predict(ELA_diff_RF, ELA_DF_Test_X)
#mae(predicted, ELA_DF_Test$school_state_diff)

#actual <- ELA_DF_Test$school_state_diff




#R2 <- 1 - (sum((actual-predicted)^2)/sum((actual-mean(actual))^2))

#R2

# mean(abs(actual-predicted))
# 
# mean(abs(actual))
# sd(abs(actual))
# sd(abs(actual-predicted))

```

```{r}

varImpPlot(ELA_diff_RF_text)

```

### Support Vector Regression

###Base SVM Model

```{r}

ELA_SVM_Train<-ELA_DF_Train%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`, `grade_level`, school_state_diff)
#Math_DF_Train_X
#Math_DF_Train$school_state_diff

svm_ela_base<-svm(school_state_diff ~., data = ELA_SVM_Train, scale = FALSE, kernel = "linear")
                     
svm_ela_base

svm_ela_pred<-predict(svm_ela_base, ELA_DF_Test_X)

 mae(svm_ela_pred, ELA_DF_Test$school_state_diff)
 
 mean(abs(ELA_DF_Test$school_state_diff))
 
  sd(abs(ELA_DF_Test$school_state_diff))
# 

```

```{r}
svmCoefs <- as.data.frame(t(coefficients(svm_ela_base)))

svmCoefs


```

### SVM w/text feature Model

```{r}

ELA_SVM_Text_Train<-ELA_DF_Train%>%
  select(`standard`, `reporting category`, `Cluster`, `text_type`, `year`, `state%`, `grade_level`, school_state_diff, poetry_factor)
#Math_DF_Train_X
#Math_DF_Train$school_state_diff

svm_ela_text<-svm(school_state_diff ~., data = ELA_SVM_Text_Train, scale = FALSE, kernel = "linear")
                     
svm_ela_text

ELA_DF_Test


svm_ela_pred<-predict(svm_ela_base, ELA_DF_Test_X)

 mae(svm_ela_pred, ELA_DF_Test$school_state_diff)
 
 mean(abs(ELA_DF_Test$school_state_diff))
 
  sd(abs(ELA_DF_Test$school_state_diff))
# 

```

```{r}

svmCoefs <- as.data.frame(t(coefficients(svm_ela_base)))

svmCoefs


```

# Results

## Math

-   Random Forest Regression Model w/ parameters from the state item reports tuned with 5-fold
    cross-validation explained 8% of the variance in the training set
    and 5% of the variance in the testing set

-   Random Forest Regression Model including 3 additional text extracted features w/ parameters tuned with 5-fold
    cross-validation explained 11% of the variance in the training set
    and 9% of the variance in the testing set

    -   The textextracted feature: `model_factor` played a more significant role than the state provided variable `type` in the variable importance plot.
    

## STE

-   Random Forest Regression Model w/ parameters from the state item reports tuned with 5-fold
    cross-validation explained 55% of the variance in the training set
    and 29% of the variance in the testing set

-   Random Forest Regression Model including 2 additional text extracted features w/ parameters tuned with 5-fold
    cross-validation explained 57% of the variance in the training set
    and 32% of the variance in the testing set

    -   The textextracted feature: `model_factor` played a more significant role than the state provided variables `pts` and `type` in the variable importance plot.
    

## ELA

The ELA model was not able to explain variance in student performance; however the model with that included text extracted features performed better than the model w/out those features.

-   Random Forest Regression Model w/ parameters from the state item reports tuned with 5-fold
    cross-validation explained -2% of the variance in the training set
    and -17% of the variance in the testing set

-   Random Forest Regression Model including 2 additional text extracted features w/ parameters tuned with 5-fold
    cross-validation explained 0% of the variance in the training set
    and -10% of the variance in the testing set

    -   The textextracted feature: `poetry_factor` and `figurative_language_factor` played a more significant role than ***most*** of the state provided variables in the variable importance plot. Notably the variable `grade_level` had the highest significance. This is interesting as the standards are the same in ELA across all grade levels. In our school there is one teacher for each grade level. This seems to suggest that in English Language Arts there is a broad inconsistency in how the teachers are addressing the standards within their courses. 

# Questions

\
At this phase, you might have more questions than answers. Document all
those questions.

1.  For each subject, I ran cross validation to tune the parameters of a Random Forest models that did not include text features and to tune the parameters of a Random Forest model that did include text features. Then I compared the %variance explained and the RMSE and MAE as well as differences in the variable importance plot. I am assuming these are the key items to discuss in my "Results" section. Is there anything else I should be adding. I wrote a few bullet points for now that I will have to flesh out in my final submission.


2.  I tried a SVM regression model; however because there are so many
    different standards, I received error messages when my test set
    contained values of the variable `standard` that weren't present in
    the training set. Is there a way to work around this? This was an
    issue for the Science and the Math tests but not for the ELA test
    since the standards are the same across all the grades. I can use a
    random forest for all of the tests without issue. Is it ok to just stick with the Random Forest Regressions?

3.  The model explained much more of the variance in the student performance in the Science Exam than the other exams. This was an interesting surprise/; is it ok to conjecture ideas as to why that could be true in my writing?

4. I have tried to comment out many of the lines of code that were "sanity checks" for me and produced tables. When I create tokens, I have been printing the summary of the tokens. This takes up a lot of space. Is this something you want to see in the paper to get a sense of the corpus, or should I comment out these summaries?
